BExIS Manual for Traitdata

### 2018-02-28 bexis_handbook/bexis_handbook_traits.Rmd
update bexis handbook for traits, include comments of Andreas, Nadja, Malte, Caterina.

---
title: "BExIS Manual for Traitdata"
author: "Florian D. Schneider"
date: "13 February 2018"
output:  pdf_document
bibliography: ../manuscript/div05.bib
csl: ../manuscript/amnat.csl
---

Functional traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.

@@

The vocabulary proposed here is meant for two-dimensional spreadsheet data. These simple table formats are what a typical research project on a particular organism group and locality produces. The vocabulary includes a core section, as well as a series of terms which are used to define the level of the reported measurement or fact (e.g. measured on individuals, or reported for an entire species), or to relate it to a broader experimental or ecosystem context. The vocabulary also contains terms for describing metadata acompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend the use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a]. ]

We assume that data are stored in a observation table, where each row is the reported measurement or fact for a single observation. At minimum, the following columns are required: 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 2. a descriptive trait name (`traitName`);  3. the scientific taxon for which the measurement or fact was obtained (`scientificName`).^[Another common format for trait data are species--trait matrices, where each row is a species and multiple trait values are stored in columns. These have to be converted into observation tables when applying the trait vocabulary.]
<!--For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers. -->

@@

To make data compatible with a consensus standard of taxon names or a look-up tables of traits, it is highly recommended to map them to globally valid terminologies by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g. GBIF Backbone Terminology)^[For *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'].

@@

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. Here, the assignment of a unique identifier per single measurement event (`measurementID`) or individual occurrence (`occurrenceID`), defines the structure these data would take in a relational database context.

The data standard provides two extensions of the vocabulary that should be used to describe these information:

- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age. This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
- The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation. It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g. variation in measurement method) or aggregated values (e.g. statistical method applied), as well as the source of reported facts (e.g. literature source or expert reference).   
- The `BiodiversityExploratories` extension provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`

@@

A comprehensive documentation of the package is found in the package website (http://EcologicalTraitData.github.io/traitdataform/), and on its Github repository (https://github.com/EcologicalTraitData/traitdataform).

@@

The key function of the package is `as.traitdata()`. This function maps column names into terms provided in the standard trait vocabulary and adds metadata as attributes to the output object. The function also will mould a species-trait-matrix into a observation table format. If data are measured on a single individual the function adds an occurrenceID to keep the reference context of individual measurements. This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:

@@

```
datasetStd <- standardize(read.csv("path/to/data.csv"),
            thesaurus = traitlist,
            taxa = "name_correct", 
            units = "mm"
            )
```

When saving your data, it is recommended to use

@@

On page 2: Most likely, the number of plots is irrelevant to your trait data, but assign whether it is about organisms from the grassland or the forest plots (or experimental manipulation), or from above or below ground system. The taxon field is of particular importance here. The field 'bioticDataTypes'  should be set to "Trait Data".

On page 3: describe the methodology underlying your trait values thoroughly. If data were extracted from literature or other database sources, declare that data are "literature data" and - if applicable - point to the column "basisOfRecord" where you should distinguish the source of your data.

On page 4: When providing keywords, you should specify which kind of traits are reported (e.g. "morphological", "behaviour", "life-history", "functional").

On page 5: Please pay attention to select the correct time format, if sampling or measuring dates are given (most likely, "yyyy-MM-dd").

On page 6: Set 'exploratory' and 'keyPlots' to 'none' if your trait data are not assigned to a single plot and site. If you created a .txt or .csv file from the R package, chose fileFormat 'ASCII' and pay attention to the field delimiter and decimal Character. Missing values created by the R package are empty fields, thus enter "".  Known missing values (i.e. values have not been measured for any reason) should not exist in the observation table format (the R package eliminates them by default). If you chose to keep rows for lacking trait measurements in the table, they should be marked as "NA". The section on 'matrix data' should be irrelevant if the data have been correctly transferred into observation tables. The variables (i.e. column names) are already provided by the template. You may add non-standard columns if necessary.

<!--Eliminate columns that are not present in your dataset. 
-->

On page 7: if the trait data are extracted from one or multiple existing databases, it is necessary to refer to those here (by name and by original paper reference). If the data have been uploaded before in a non-standard format, please provide the dataset ID.

>>>>

---
title: "BExIS Manual for Traitdata"
author: "Florian D. Schneider (florian.dirk.schneider@gmail.com)"
date: "28 February 2018 (v0.2)"
output:  
  pdf_document:
    includes:
      in_header: header.tex
    toc: yes
    fig_caption: true
fontsize: 10pt
geometry: margin=1.5in
bibliography: ../manuscript/div05.bib
csl: ../manuscript/amnat.csl
---

# Ecological traits

Ecological traits have become a core concept of ecological research and have been assessed for many organism groups within the Biodiversity Exploratories project.

@@

The vocabulary proposed here is designed for two-dimensional spreadsheet data. These simple table formats are what a typical research project on a particular organism group and locality produces. The vocabulary includes a core section, as well as a series of terms which are used to define the level of the reported measurement or fact (e.g. measured on individuals, or reported for an entire species), or to relate it to a broader experimental or ecosystem context. The vocabulary also contains terms for describing metadata accompanying the trait dataset.^[Two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. For datasets containing multi-layered information on observations, traits, taxa and environmental context, we recommend to use relational datatabase structures composed of multiple spreadsheets linked via identifiers, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a]. ]

We assume that data are stored in an observation table, where each row is the reported measurement or fact for a single observation. At minimum, the following columns are required (Fig. 1): 1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 2. a descriptive trait name (`traitName`);  3. the scientific taxon for which the measurement or fact was obtained (`scientificName`).^[Another common format for trait data are species--trait matrices, where each row is a species and multiple trait values are stored in columns. These have to be converted into observation tables when applying the trait vocabulary.]
<!--For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers. -->

![Elements of the original data set (species $\times$ trait matrix) are transferred into a observation table format with additional columns for trait and taxon standardization. Further information on the level of measurements, occurrences (i.e. individual specimens) and sampling locations can be added via the extended vocabulary.](figures.PNG)

@@

To make data compatible with a consensus standard of taxon names or a look-up tables of traits, it is highly recommended to map the dataset-specific names provided for taxa (`scientificName`) and traits (`traitName`) onto globally valid terminologies by adding uniform ressource identifiers ([URIs](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier)) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by human-readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively; e.g. GBIF Backbone Terminology; see Fig. 1)^[For *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'].

@@

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. Here, the assignment of a unique identifier per single measurement event (`measurementID`) or individual occurrence (`occurrenceID`), defines the structure these data would take in a relational database context (Fig. 1).

The data standard provides two extensions of the vocabulary that should be used to describe this information:

- The `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation. It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g. variation in measurement method) or aggregated values (e.g. statistical method applied), as well as the source of reported facts (e.g. literature source or expert reference).
- The `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age. This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.
- The `BiodiversityExploratories` extension  (http://ecologicaltraitdata.github.io/ETS/#extension-biodiversity-exploratories)  provides terms to relate the trait measurement to plots and sites in the exploratories framework, including `ExploratoriesPlotID`, `Exploratory` and `ExploType`.

@@

A comprehensive documentation of the package is found on the package website (http://EcologicalTraitData.github.io/traitdataform/), and on its Github repository (https://github.com/EcologicalTraitData/traitdataform).

@@

The key function of the package is `as.traitdata()`. This function maps column names into terms provided in the standard trait vocabulary and adds metadata as attributes to the output object. The function also will mould a species-trait-matrix into an observation table format. If data are measured on a single individual the function adds an occurrenceID to keep the reference context of individual measurements. This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:

@@

```
datasetStd <- standardize(read.csv("path/to/data.csv"),
            thesaurus = traitlist,
            taxa = "name_correct", 
            units = "mm",
            bexis = TRUE
            )

write.csv(datasetStd, file = "filename.csv", row.names = FALSE)

This creates a clean text file (i.e. it omits the row names when producing a comma separated file with points as decimal delimiter) ready for upload on BExIS.

@@

**On page 2:** Please assign the number of plots the data are related to (leave empty for data not assessed from exploratory specimens), and whether it is about organisms from the grassland or the forest plots (or experimental manipulation), or from above or below ground system. The taxon field is of particular importance here. The field 'bioticDataTypes'  should be set to "Trait Data".

**On page 3:** describe the methodology underlying your trait values thoroughly. If data were extracted from literature or other database sources, declare that data are "literature data" and - if applicable - point to the column "basisOfRecord" where you should distinguish the source of your data.

**On page 4:** When providing keywords, you should specify which kind of traits are reported (e.g. "morphological", "behaviour", "life-history", "functional").

**On page 5:** Please pay attention to select the correct time format, if sampling or measuring dates are given (most likely, "yyyy-MM-dd").

**On page 6:** Set 'exploratory' and 'keyPlots' to 'none' if your trait data has no column with specific exploratories information, otherwise enter the column that contains the plot ID (by default "ExploratoriesPlotID"). If you created a .txt or .csv file from the R package, choose fileFormat 'ASCII' and double check for the correct field delimiter (usually ",") and decimal separator (usually "."). Structural missing values for empty columns created by the R package are empty fields, thus enter "". 
Known missing values (i.e. trait values which have not been measured for any reason) should not exist in the observation table format (the R package eliminates them by default, if they were correctly marked as NA in the original data). If you choose to keep rows for lacking trait measurements in the table, they should be marked as "NA". The section on 'matrix data' can be ignored since the data have been transferred into observation table format. The variables (i.e. column names) are already provided by the template. You may add non-standard columns if necessary.

**On page 7:** if the trait data are extracted from one or multiple existing databases, it is necessary to refer to those here (by name and by original paper reference). If the data have been uploaded before in a non-standard format, please provide the dataset ID.

# Assistance and Feedback

This project is work in progress. If something doesn't work as expected, please get in touch and report problems:

- For troubleshooting during file upload, please contact BExIS Helpdesk (mail@?.de). For issues with the definitions of the vocabulary, please get involved in the discussion on Github (https://github.com/EcologicalTraitData/ETS/issues). 
- For comments on this manual, please contact us directly (florian.dirk.schneider@gmail.com). 
- For bug reports or feature requests of the R-package, please file an issue on Github (https://github.com/EcologicalTraitData/traitdataform/issues).
